Word-Referent Identification Under Multimodal Uncertainty

### 2018-04-26 Writeup/ms.Rmd
modif ms

 Identifying a spoken word in a referential context requires both the ability to process and integrate multimodal input and the ability to reason under uncertainty. How do these tasks interact with one another? We introduce a task that allows us to examine how adults identify words under joint uncertainty in the auditory and visual modalities. We propose an ideal observer model which provides an account of how auditory and visual cues are combined optimally. Model predictions are tested in three experiments where word recognition is made under two kinds of uncertainty: category ambiguity and/or distorting noise. In all cases, the optimal model explains much of the variance in human mean judgments. In particular, when the signal is not distorted with noise, participants weight the auditory and visual cues optimally, that is, according to the relative reliability of each modality. But when one modality has noise added to it, human perceivers systematically prefer the unperturbed modality to a greater extent than the optimal model does. The study provides a formal framework which helps us understand precisely how word form and word meaning interact in word recognition under uncertainty. Moreover it offers a first step towards a model that accounts for form-meaning synergy in early word learning.

@@

The task of matching the sound to the corresponding visual object has been extensively studied in the developmental literature since it is considered to be an crucial instance of early word learning. For example, many studies focused on how children succeed in this task despite high referential ambiguity [@smith08; @pinker1989; @medina2011; @vouloumanos2008; @yurovsky2015; @suanda2014; @vlach2013]. However, even when they know the exact meanings of the words, listeners (both children and adults) often face the task of recognizing which word the speaker has uttered, especially under noisy circumstances. The purpose of the current study is to explore the special case of word recognition under uncertainty when adult observers have access to multimodal cues from the speech and the referent.

One rigorous way to approach this question is through conducting an *ideal observer* analysis. This research strategy provides a characterization of the task/goal and shows what the optimal performance should be under this characterization.^[It is, thus, a general instance of the rational approach to cognition [@anderson90]. It can also be seen as an instance of Marr's computational level of analysis.] When there is uncertainty in the input, the ideal observer performs an optimal probabilistic inference. For example, in order to recognize an ambiguous linguistic input, the model uses all available probabilistic knowledge in order to maximize the accuracy of this recognition. The ideal observer model can be seen as a theoretical upper limit on performance. It is not so much a realistic model of human performance, as much as a baseline against which human performance can be compared [@Geisler2003; @rahnev2018]. When there is a deviation from the ideal, it can reveal extra constraints on human cognition, such as limitations on the working memory or attentional resources.  The ideal observer analysis has had a tremendous impact not only on speech related research [@Norris08; @clayard08; @feldman2009; @kleinschmidt2015], but also on many other disciplines in the cognitive sciences [for reviews, see @chater06; @Knill04; @tenenbaum11]

Some of these ideal-observer-based studies are closely related to the question we are addressing in the current work. For instance, @clayard08 simulated auditory uncertainty by manipulating the probability distribution of a cue (Voice Onset Time) that differentiated similar words (e.g., "beach" and "peach"). They found that humans were sensitive to these probabilistic cues and their judgments closely reflected the optimal predictions. In another work, @feldman2009 studied the perceptual magnet effect, which is a phenomenon that involves reduced discriminability near prototypical sounds in the native language [@kuhl1991]. They showed that this effect can be explained as the consequence of optimally solving the problem of perception under uncertainty.

@@

On the face of it, the question of combining information from the sound and the visual referent might seem similar to that of audio-visual speech integration. Nevertheless, there are at least two fundamental differences between these two cases, and both can influence the way the auditory and visual cues are combined: First, in the case of audio-visual speech, both modalities offer information about the same underlying speech category. They may differ only in terms of their informational reliability. In a referential context, however, the auditory and visual modalities play different roles in the referential process---in addition to possible differences in informational reliability. Indeed, the auditory input represents the *symbol* whereas the visual input represents the *meaning*. It has been suggested that because of its referential property, speech is a privileged signal for humans, starting in infancy [see @vouloumanos2014 for a review].^[There is a debate as to whether speech is privileged for children and adults for the same reasons. Whereas some researchers suggest that speech is privileged for both children and adults because of its ability to refer [e.g., @waxman1995], others suggest that speech might *not* have a referential status from the start. Rather, speech might be preferred by children only because of a low level auditory ``overshadowing'' [e.g., @sloutsky2003].] Thus, in a referential context, it is possible that listeners do not treat the auditory and visual modalities as equivalent sources of information. Instead, there could be a sub-optimal bias for the auditory modality beyond what is expected from informational reliability alone.

@@

We investigate how people combine cues from the auditory and the visual modality to recognize words in a referential context. In particular, we study how this combination is performed under various degrees of uncertainty in both the auditory and the visual modality. Imagine, for example, that someone is uncertain whether they heard "pea" or "bee", does this uncertainty make them rely more on the referent (e.g., the object being pointed at)? Vice versa, if they are not sure if they saw a bee or a fly, does it make them rely more on the sound? More importantly, when input in both modalities is uncertain to varying degrees, do they weight each modality according to its relative reliability (which is the optimal strategy), or do they over-rely on a particular modality (which is a sub-optimal strategy)?

@@

We introduce a new task that tests word recognition in a referential context.  We use two visual categories (cat and dog) and two auditory categories (/b/ and /d/ embedded in the minimal pair /aba/-/ada/). For each participant, an arbitrary pairing is set between the auditory and the visual categories, leading to two audio-visual word categories (e.g., dog-/aba/, cat-/ada/). In each trial, participants are presented with an audio-visual target (the prototype of the target category), immediately followed by an audio-visual test stimulus (Figure\ \@ref(fig:task)). The test stimulus may differ from the target in both the auditory and the visual components.  After these two presentations, participants press "same" or "different."

@@

This paradigm is adapted from a previous task [@sloutsky2003], which has been used with both children and adults to probe audio-visual encoding [see @robinson2010 for a review]. In the testing phase of the original task, participants are asked whether or not the two audio-visual presentations are *identical*. In the current study, we are interested, rather, in the categorization, i.e., determining whether or not two similar tokens are members of the same phonological/semantic category. Therefore, testing in our task is category-based: Participants are asked to press "same" if they think the second item (the test) belongs to the same category as the first (target) (e.g.,  dog-/aba/), even if there is a slight difference in the sound, in the referent, or in both. They are instructed to press "different" only if they think that the second stimulus was an instance of the other category (cat-/ada/). The task also includes trials where pictures are hidden (audio-only) or where sounds are muted (visual-only). These unimodal trials provide us with the participants' evaluation of the probabilistic information present in the auditory and visual categories. As we shall see, these unimodal distributions are used as inputs to the optimal cue combination model.

@@

We construct an ideal observer model that combines probabilistic information from the auditory and visual modalities. In contrast to the  model used in most research on multisensory integration [e.g., @ernst02]---which typically studies continuous stimuli (e.g., size, location)---the probabilistic information in our case cannot be characterized with *sensory noise*, only.  Indeed, our task involves responses over categorical variables (phonemes and concepts), and therefore, the optimal model should take into account, not only the noise variability around an individual perceptual estimate, but also its *categorical variability*, i.e., the uncertainty related to whether this perceptual estimate belongs to a given category [see also @Bankieris17; @bejjanki2011]. In what follows, we describe a model that accounts for both type of variability. First, we describe the model in the simplified case of categorical variability only. Second, we augment this simplified model to account for sensory noise.

@@

\noindent This assumption says that, given a word-object mapping, e.g., $W=$("cat"-CAT), variation in the way "cat" is pronounced does not correlate with changes in any visual property of the object CAT, which is a valid assumption.\footnote{Note that this assumptions is more adequate in the case of arbitrary associations such as ours, and less so in the case of redundant association such as audio-visual speech. In the latter, variation in the pronunciation is expected to correlate, at least to some extent, with lip movements.}

@@

The parameter $b$ represents the differential between the categories' prior probabilities. However, since the identity of word categories is randomized across participants, $b$ measures, rather, a response bias to "same" if $b > 0$, and a response bias to "different" if $b < 0$. We expect a general bias towards answering "different" because of the categorical nature of our same-different task: When two items are ambiguous but perceptually different, participants might have a slight preference for "different" over "same". As for the means, their values are fixed, and they correspond to the most typical tokens in our stimuli. Finally, observations from each modality ($a$ and $v$) are weighted in Equation 1 according to their reliability: $$\beta_a \propto \frac{1}{\sigma^2_{A}}$$ $$\beta_v \propto \frac{1}{\sigma^2_{V}}.$$

@@

So far, we only accounted for categorical variability. For instance, if the speaker generates a target production $a_t$ from an auditory category
$p(a_t | A) \sim N(\mu_{A}, \sigma^2_{A})$, the ideal model assumes that it has direct access to this production token (i.e., $a=a_t$), and that all uncertainty is about the category membership of this token. However, we might also want to account for internal noise in the brain and/or external noise in the environment.  For example, the observer might not have access to the exact produced target, but only to the target perturbed by noise. If we assume this noise to be normally distributed, that is,  $p(a | a_t) \sim N(a_t, \sigma^2_{N_A})$, then integrating over $a_t$ leads to the following simple expression:
$$ p(a | A) \sim N(\mu_{A}, \sigma^2_{A}+\sigma^2_{N_A})$$
Similarly, in the case of sensory noise in the visual modality, we get
$$ p(a | V) \sim N(\mu_{V}, \sigma^2_{V}+\sigma^2_{N_V})$$
Finally, using exactly the same derivation as above, we end up with the following multimodal weighting scheme in the optimal combination model (Equation 1) which takes into account both categorical and sensory variability:

@@

Equation 1 provides the optimal model's predictions for how probabilities that characterize uncertainty in the auditory and the visual modalities can be combined to make categorical decisions. Parameters' estimates of the probability distributions in each modality are derived by fitting unimodal posteriors to the participants' responses in the unimodal conditions, i.e., the condition where only the sounds are heard  or only the pictures are seen (Figure\ \@ref(fig:task)).\footnote{Further technical detail about model fitting in the unimodal conditions will be given in the method section of Experiment 1} Using these derived parameters, the optimal model makes predictions about responses in the bimodal condition where participants both hear the sounds and see the pictures.

@@

Comparison of the optimal and the descriptive models allows us, not only to quantify how much people deviate from optimality, but also to understand precisely the nature of this deviation. Let $\sigma^2_{A}$ and $\sigma^2_{V}$ be the values of the variances used in the optimal model (derived from the unimodal conditions), and $\sigma^2_{Ab}$ and $\sigma^2_{Vb}$ be the values observed through the descriptive model in the bimodal condition. Deviation from optimality is measured in two ways. First, we measure the change in the values of the variance specific to each modality, that is, how $\sigma^2_{A}$ compares to $\sigma^2_{Ab}$, and how $\sigma^2_{V}$ compares to $\sigma^2_{Vb}$. Second, we measure changes in the proportion of the visual and auditory variances, i.e., we examine how $\frac{\sigma^2_{A}}{\sigma^2_{V}}$ compares to $\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}}$. The first measure allows us to test if response precision changes for each modality when we move from the unimodal to the bimodal conditions. The second allows us to test the extent to which the weighting scheme follows the prediction of the optimal model.  The reason we used the proportion of the variances as a measure of cross-modal weighting is because this proportion corresponds to the slope\footnote{Or more precisely the absolute value of the slope} of the decision threshold in the audio-visual space (Figure\ \@ref(fig:model)). The decision threshold is defined as the set of values in this audio-visual space along which the posterior is equal to 0.5. Formally speaking, the decision threshold has the following form:

@@

In this Experiment, we test the predictions of the model in the case where uncertainty is due to categorical variability only (i.e., ambiguity in terms of category membership). We do not add any external noise to the background and we assume that internal sensory noise is negligible compared to categorical variability ($\sigma^2_{A} \gg \sigma^2_{N_A}$, and $\sigma^2_{V} \gg \sigma^2_{N_V}$). Thus, we use the following cue weighting scheme:

@@

The first part of the experiment trained participants using only the prototype pictures and the prototype sounds (12 trials, 4 each from the bimodal, audio-only, and visual-only conditions). After completing training, we instructed participants on the structure of the task and encouraged them to base their answers on both the sounds and the pictures (in the bimodal condition). There were a total of 25 possible combinations in the bimodal condition, and 5 in each of the unimodal conditions. Each participant saw each possible trial twice, for a total of 70 trials/participant. Trials were blocked by condition and blocks were presented in random order. The experiment lasted around 15 minutes.

@@

### plot unimodal data + fit functions 
all_plot <- ggplot(exp_uni_data, 
       aes(x = distance, y = mean)) + 
  geom_point()+
  geom_pointrange(aes(ymin = lower, ymax = upper), 
                  position = position_dodge(width = .1)) + 
  #geom_line(data=uniV,aes(x=xV, y=yV))+
  geom_line(data=all_uni_model, aes(x=distance, y=prediction), col='black')+
  xlab("Distance") +ylab("Prob. different")+
  scale_y_continuous(limits = c(0, 1))+#theme(aspect.ratio = 0.7)+
  theme_few()+
  theme(aspect.ratio = 0.7) + facet_grid(Experiment ~ Condition)
  #stat_function(fun = Logistic_v, colour="red"))

@@

### Model fitting details
#### Unimodal condition
Remember that data in this conditions allow us to derive the variances of both the auditory and the visual categories, and that these variances are used to make predictions about bimodal data (in the visual and auditory baselines as well as in the optimal model). These individual variances were derived as follows (we explain the derivation for the auditory-only case, but the same applies for the visual-only case). We use the same Bayesian reasoning as we did in the derivation of the bimodal model: When presented with an audio instance $a$, the probability of choosing the sound category 2 (that is, to answer "different") is the posterior probability of this category $p(A_2|a)$. If we assume that both sound categories have equal variances, the posterior probability reduces to:

@@

Although the paradigm is within-subjects, we did not have enough statistical power to fit a different model for each individual participant. Instead, models were constructed with data collapsed across all participants. That being said, the distribution of responses from individual participants will also be analyzed. The fit was done with a nonlinear least squares regression using the NLS package in R [@bates88]. We computed the values of the parameters, as well as their 95% confidence intervals, through non-parametric bootstrap (using 10000 iterations).

@@

Average categorization judgments and best fits are shown in Figure\ \@ref(fig:unimodal). The categorization function of the auditory condition was slightly steeper than that of the visual condition, meaning that participants perceived the sound tokens slightly more categorically and whih higher certainty than they did with the visual tokens.  For the auditory modality, we obtained the following values:^[all CIs in the paper are 95% confidence intervals.] $b_A=$ `r audBias_val`  [`r audBias_ci2`, `r audBias_ci1`] and $\sigma^2_A=$ `r audVar_val` [`r audVar_ci1`, `r audVar_ci2`]. For the visual modality, we obtained $b_V=$ `r visBias_val` [`r visBias_ci2`, `r visBias_ci1`] and $\sigma^2_V=$ `r visVar_val` [`r visVar_ci1`, `r visVar_ci2`].

```{r unimodal, out.width = "\\textwidth", fig.pos = "!h", fig.cap = "Human responses in the unimodal conditions. Points represent the proportion of `different' to `same' responses in the auditory-only condition (left), and visual-only condition (right). Error bars are 95\\% confidence intervals. Solid lines represent best fits."}

@@

correlation_plot <- ggplot(models_all, 
       aes(x = pred, y = joint, col = factor(concept_dist), 
           shape = factor(sound_dist))) + 
  geom_point()+
  scale_colour_solarized()+
 #geom_pointrange(aes(ymin = summary_ci_lower, ymax = summary_ci_upper), 
  #                position = position_dodge(width = .1), size=0.2) +

@@

#### Normative models
Figure\ \@ref(fig:bimodal) compares the predictions of the normative models against human responses. The visual, auditory and optimal model explained, respectively, `r 100*R2_visual_1`%, `r 100*R2_auditory_1`%, and `r 100*R2_optimal_1`% of total variance in mean responses.

@@

```{r bimodal, out.width = "\\textwidth", fig.pos = "!h", fig.cap = "Human responses vs. models' predictions in the bimodal condition. Shape represents auditory distance from the target, and color represents visual distance from the target."}

@@

## Discussion
Overall, we found that the optimal model explained much of the variance in the mean judgments, and largely more than what can be explained with the auditory or the visual models alone. Moreover, the high value of the coefficient of determination in the optimal model ($r^2$=`r R2_optimal_1`) suggests that the population was near-optimal. However, we see in Figure\ \@ref(fig:bimodal) that the mean responses deviated systematically from the optimal prediction in that they were slightly pulled toward chance (i.e., the probability 0.5). This fact is due to the increase in the value of the variance associated with each modality. Note however that, despite this increase in randomness, our analysis of modality preference showed that the relative values of these variances were not different (Figure\ \@ref(fig:bias)), meaning that there was no evidence for a modality preference. Thus, 1) There was a simultaneous increase in the values of the auditory and visual variances in the bimodal condition compared to the unimodal condition, meaning that the bimodal input lead to an increase in response randomness, and 2) this increased randomness did not affect the relative weighting of both modalities, i.e., the population was weighting modalities according to the relative reliability predicted by the optimal model. This situation corresponds to the first case of sub-optimally described in Figure\ \@ref(fig:subOptim).

@@

```{r echo=FALSE, individual, out.width = "\\textwidth", fig.pos = "!h",fig.cap = "Histograms of the individual values of the visual variance relative to the auditory variance in Experiment 1. Light color represents the values derived from each individual participant, and dark color represents simulated individual responses sampled from the descriptive model."}

@@

In Experiment 1, we tested word recognition when there was multimodal uncertainty in terms of category membership only. In real life, however, tokens can undergo distortions due to noisy factors in the environment (e.g., car noise in the background, blurry vision in a foggy weather,..). In Experiment 2 and 3, we explore this additional level of uncertainty.

>>>>

 Identifying a spoken word in a referential context requires both the ability to integrate multimodal input and the ability to reason under uncertainty. How do these tasks interact with one another? We introduce a task that allows us to examine how adults identify words under joint uncertainty in the auditory and visual modalities. We propose an ideal observer model which provides an account of how auditory and visual cues are combined optimally. Model predictions are tested in three experiments where word recognition is made under two kinds of uncertainty: category ambiguity and/or distorting noise. In all cases, the optimal model explains much of the variance in human mean judgments. In particular, when the signal is not distorted with noise, participants weight the auditory and visual cues optimally, that is, according to the relative reliability of each modality. But when one modality has noise added to it, human perceivers systematically prefer the unperturbed modality to a greater extent than the optimal model does. The study provides a formal framework which helps us understand precisely how word form and word meaning interact in word recognition under uncertainty. Moreover it offers a first step towards a model that accounts for form-meaning synergy in early word learning.

@@

The task of matching the sound to the corresponding visual object has been extensively studied in the developmental literature since it is considered to be an crucial instance of early word learning. For example, many studies focused on how children might succeed in this task despite high referential ambiguity [@smith08; @pinker1989; @medina2011; @vouloumanos2008; @yurovsky2015; @suanda2014; @vlach2013]. However, even when they know the exact meanings of the words, listeners (both children and adults) often face the task of recognizing which word the speaker utters, especially under noisy circumstances [@mattys12; @peelle2018]. The purpose of the current study is to explore novel word recognition by adults under multimodal uncertainty. We focus on the special case where people have access to multimodal cues from the auditory speech and the visual referent.

One rigorous way to approach this question is through conducting an *ideal observer* analysis. This research strategy provides a characterization of the task/goal and shows what the optimal performance should be under this characterization.^[It is, thus, a general instance of the rational approach to cognition [@anderson90]. It can also be seen as an instance of Marr's computational level of analysis [@marr1982].] When there is uncertainty in the input, the ideal observer performs an optimal probabilistic inference. For example, in order to recognize an ambiguous linguistic input, the model uses all available probabilistic knowledge in order to maximize the accuracy of this recognition. The ideal observer model can be seen as a theoretical upper limit on performance. It is not so much a realistic model of human performance, as much as a baseline against which human performance can be compared [@Geisler2003; @rahnev2018]. When there is a deviation from the ideal, it can reveal extra constraints on human cognition, such as limitations on the working memory or attentional resources.  The ideal observer analysis has had a tremendous impact not only on speech related research [@Norris08; @clayard08; @feldman2009; @kleinschmidt2015], but also on many other disciplines in the cognitive sciences [for reviews, see @chater06; @Knill04; @tenenbaum11]

Some of these ideal-observer-based studies are closely related to the question we are addressing in the current work. For instance, @clayard08 simulated auditory uncertainty by manipulating the probability distribution of a cue (Voice Onset Time) that differentiated similar words (e.g., "beach" and "peach"). They found that humans were sensitive to these probabilistic cues and their judgments closely reflected the optimal predictions. In another work, @feldman2009 studied the perceptual magnet effect, a phenomenon that involves reduced discriminability near prototypical sounds in the native language [@kuhl1991]. They showed that this effect can be explained as the consequence of optimally solving the problem of perception under uncertainty.

@@

On the face of it, the question of combining information from the sound and the visual referent might seem similar to that of audio-visual speech integration. Nevertheless, there are at least two fundamental differences between these two cases, and both can influence the way the auditory and visual cues are combined: First, in the case of audio-visual speech, both modalities offer information about the same underlying speech category. They may differ only in terms of their informational reliability. In a referential context, however, the auditory and visual modalities play different roles in the referential process---in addition to possible differences in informational reliability. Indeed, the auditory input represents the *symbol* whereas the visual input represents the *meaning*. It has been shown that speech has a privileged status compared to other sensory stimuli [@vouloumanos2014; @waxman2009; @waxman1995; @edmiston2015; @lupyan2012], and that this privilege is indeed related to the speech's ability to refer [@waxman2009].^[There is, however, a debate as to whether speech is privileged for children and adults for similar reasons. Whereas some researchers suggest that speech is privileged for both children and adults because of its ability to refer [e.g., @waxman2009], others suggest that speech might *not* have a referential status from the start. Rather, speech might be preferred by children only because of a low level auditory ``overshadowing'' [e.g., @sloutsky2003].] Thus, in a referential context, it is possible that listeners do not treat the auditory and visual modalities as equivalent sources of information. Instead, there could be a sub-optimal bias for the auditory modality beyond what is expected from informational reliability alone.

@@

We investigate how people combine cues from the auditory and the visual modality to recognize words in a referential context. In particular, we study how this combination is performed under various degrees of uncertainty in both the auditory and the visual modality. Imagine, for example, that someone is uncertain whether they heard "pea" or "bee", does this uncertainty make them rely more on the referent (e.g., the object being pointed at)? Vice versa, if they are not sure if they saw a bee or a fly, does it make them rely more on the sound? More importantly, when input in both modalities is uncertain to varying degrees, do they weight each modality according to its relative reliability, which is the optimal strategy, or do they over-rely on a particular modality, which is a sub-optimal strategy?

@@

We introduce a task adapted from @sloutsky2003. The original task has been used with both children and adults to probe audio-visual encoding [see @robinson2010 for a review]. Here we use a slightly different version to test word recognition in a referential context.  We use two visual categories (cat and dog) and two auditory categories (/b/ and /d/ embedded in the minimal pair /aba/-/ada/). For each participant, an arbitrary pairing is set between the auditory and the visual categories, leading to two audio-visual word categories (e.g., dog-/aba/, cat-/ada/). In each trial, participants are presented with an audio-visual target (the prototype of the target category), immediately followed by an audio-visual test stimulus (Figure\ \@ref(fig:task)). The test stimulus may differ from the target in both the auditory and the visual components.  After these two presentations, participants press "same" or "different."

@@

In the testing phase of the original task [@sloutsky2003], participants are asked whether or not the two audio-visual presentations are *identical*. In the current study, we are interested, rather, in the categorization, i.e., determining whether or not two similar tokens are members of the same phonological/semantic category. Therefore, testing in our task is category-based: Participants are asked to press "same" if they think the second item (the test) belongs to the same category as the first (target) (e.g.,  dog-/aba/), even if there is a slight difference in the sound, in the referent, or in both. They are instructed to press "different" only if they think that the second stimulus was an instance of the other category (cat-/ada/). The task also includes trials where pictures are hidden (audio-only) or where sounds are muted (visual-only). These unimodal trials provide us with the participants' evaluation of the probabilistic information present in the auditory and visual categories. As we shall see, these unimodal distributions are used as inputs to the optimal cue combination model.

@@

We construct an ideal observer model that combines probabilistic information from the auditory and visual modalities. In contrast to the  model used in most research on multisensory integration [e.g., @ernst02]---which typically studies continuous stimuli (e.g., size, location)---the probabilistic information in our case cannot be characterized with *sensory noise*, only.  Indeed, our task involves responses over categorical variables (phonemes and concepts), and therefore, the optimal model should take into account, not only the noise variability around an individual perceptual estimate, but also its *categorical variability*, i.e., the uncertainty related to whether this perceptual estimate belongs to a given category [see also @Bankieris17; @bejjanki2011]. In what follows, we describe a model that accounts for both types of variability. First, we describe the model in the simplified case of categorical variability only. Second, we augment this simplified model to account for sensory noise.

@@

\noindent This assumption says that, given a word-object mapping, e.g., $W=$("cat"-CAT), variation in the way "cat" is pronounced does not correlate with changes in any visual property of the object CAT, which is a valid assumption in the context of our task.\footnote{Note that this assumptions is more adequate in the case of arbitrary associations such as ours, and less so in the case of redundant association such as audio-visual speech. In the latter, variation in the pronunciation is expected to correlate, at least to some extent, with lip movements.}

@@

The parameter $b$ represents the differential between the categories' prior probabilities. However, since the identity of word categories is randomized across participants, $b$ measures, rather, a response bias to "same" if $b > 0$, and a response bias to "different" if $b < 0$. We expect a general bias towards answering "different" because of the categorical nature of our same-different task: When two items are ambiguous but perceptually different, participants might have a slight preference for "different" over "same". As for the means, their values are fixed, and they correspond to the most typical tokens in our stimuli. Finally, observations from each modality ($a$ and $v$) are weighted in Equation 1 according to their reliability (that is, according to the *inverse* of their variance): $$\beta_a \propto \frac{1}{\sigma^2_{A}}$$ $$\beta_v \propto \frac{1}{\sigma^2_{V}}.$$

@@

So far, we only accounted for categorical variability. For instance, if the speaker generates a target production $a_t$ from an auditory category
$p(a_t | A) \sim N(\mu_{A}, \sigma^2_{A})$, the ideal model assumes that it has direct access to this production token (i.e., $a=a_t$), and that all uncertainty is about the category membership of this token. However, we might also want to account for internal noise in the brain and/or external noise in the environment.  For example, the observer might not have access to the exact produced target, but only to the target perturbed by noise. If we assume this noise to be normally distributed, that is,  $p(a | a_t) \sim N(a_t, \sigma^2_{N_A})$, then integrating over $a_t$ leads to this new expression of the probability distribution:
$$ p(a | A) \sim N(\mu_{A}, \sigma^2_{A}+\sigma^2_{N_A})$$
Similarly, in the case of sensory noise in the visual modality, we get:
$$ p(a | V) \sim N(\mu_{V}, \sigma^2_{V}+\sigma^2_{N_V})$$
Finally, using exactly the same derivation as above, we end up with the following multimodal weighting scheme in the optimal combination model (Equation 1) which takes into account both categorical and sensory variability:

@@

Equation 1 provides the optimal model's predictions for how probabilities that characterize uncertainty in the auditory and the visual modalities can be combined to make categorical decisions. Parameter estimates of the probability distributions in each modality are derived by fitting unimodal posteriors to the participants' responses in the unimodal conditions, i.e., the condition where only the sounds are heard  or only the pictures are seen (Figure\ \@ref(fig:task)).\footnote{Further technical detail about model fitting in the unimodal conditions will be given in the method section of Experiment 1} Using these derived parameters, the optimal model makes predictions about responses in the bimodal (i.e., audio-visual) condition where participants both hear the sounds and see the pictures.

@@

Comparison of the optimal and the descriptive models allows us, not only to quantify how much people deviate from optimality, but also to understand precisely the nature of this deviation. Let $\sigma^2_{A}$ and $\sigma^2_{V}$ be the values of the variances used in the optimal model (derived from the unimodal conditions), and $\sigma^2_{Ab}$ and $\sigma^2_{Vb}$ be the values observed through the descriptive model in the bimodal condition. Deviation from optimality is measured in two ways. First, we measure the change in the values of the variance specific to each modality, that is, how $\sigma^2_{A}$ compares to $\sigma^2_{Ab}$, and how $\sigma^2_{V}$ compares to $\sigma^2_{Vb}$. Second, we measure changes in the proportion of the visual and auditory variances, i.e., we examine how $\frac{\sigma^2_{A}}{\sigma^2_{V}}$ compares to $\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}}$. The first measure allows us to test if response precision changes for each modality when we move from the unimodal to the bimodal conditions. The second allows us to test the extent to which the weighting scheme follows the prediction of the optimal model.  The reason we used the proportion of the variances as a measure of cross-modal weighting is because this proportion corresponds to the slope\footnote{Or more precisely the absolute value of the slope.} of the decision threshold in the audio-visual space (Figure\ \@ref(fig:model)). The decision threshold is defined as the set of values in this audio-visual space along which the posterior is equal to 0.5. Formally speaking, the decision threshold has the following form:

@@

In this Experiment, we test the predictions of the model in the case where uncertainty is due to categorical variability only (i.e., ambiguity in terms of category membership). We do not add any external noise to the background and we assume that internal sensory noise is negligible compared to categorical variability ($\sigma^2_{A} \gg \sigma^2_{N_A}$ and $\sigma^2_{V} \gg \sigma^2_{N_V}$). Thus, we use the following cue weighting scheme:

@@

The first part of the experiment trained participants using only the prototype pictures and the prototype sounds (12 trials, 4 each from the bimodal, audio-only, and visual-only conditions). After completing training, we instructed participants on the structure of the task and encouraged them to base their answers on both the sounds and the pictures (in the bimodal condition). There were a total of 25 possible combinations in the bimodal condition, and 5 in each of the unimodal conditions. Each participant saw each possible trial twice, for a total of 70 trials/participant. Trials were blocked by condition and blocks were presented in random order. The experiment lasted around 15 minutes.^[The experiment can be accessed and played from the github repository: https://github.com/afourtassi/WordRec/]

@@

### plot unimodal data + fit functions 
all_plot <- ggplot(exp_uni_data, 
       aes(x = distance, y = mean)) + 
  #geom_point()+
  geom_pointrange(aes(ymin = lower, ymax = upper), 
                  position = position_dodge(width = .1), size = 0.4, fatten = 2) + 
  #geom_line(data=uniV,aes(x=xV, y=yV))+
  geom_line(data=all_uni_model, aes(x=distance, y=prediction), col='black')+
  xlab("Distance") +ylab("Prob. different")+
  scale_y_continuous(limits = c(0, 1))+#theme(aspect.ratio = 0.7)+
  theme_few()+
  theme(aspect.ratio = 0.7) + facet_grid(Experiment ~ Condition)
  #stat_function(fun = Logistic_v, colour="red"))

@@

### Model fitting details
#### Unimodal conditions
Remember that data in these conditions allows us to derive the variances of both the auditory and the visual categories, and that these variances are used to make predictions about bimodal data (in the visual and auditory baselines as well as in the optimal model). These individual variances were derived as follows (we explain the derivation for the auditory-only case, but the same applies for the visual-only case). We use the same Bayesian reasoning as we did in the derivation of the bimodal model: When presented with an audio instance $a$, the probability of choosing the sound category 2 (that is, to answer "different") is the posterior probability of this category $p(A_2|a)$. If we assume that both sound categories have equal variances, the posterior probability reduces to:

@@

Although the paradigm is within-subjects, we did not have enough statistical power to fit a different model for each individual participant.^[We had a relatively high number of trials spanning all possible audio-visual matchings. Getting enough data points per trial per participant would have required running the online experimental for a much longer time (more than an hour). This could have increased significantly the dropout rate and possibly affected the quality of the data collected online.] Instead, models were constructed with data collapsed across all participants. That being said, we will also analyze the distribution of individual responses. The fit was done with a nonlinear least squares regression using the NLS package in R [@bates88]. We computed the values of the parameters, as well as their 95% confidence intervals, through non-parametric bootstrap (using 10000 iterations).

@@

Average categorization judgments and best fits are shown in Figure\ \@ref(fig:unimodal). The categorization function of the auditory condition was slightly steeper than that of the visual condition, meaning that participants perceived the sound tokens slightly more categorically and whih higher certainty than they did with the visual tokens.  For the auditory modality, we obtained the following values: $b_A=$ `r audBias_val`  [`r audBias_ci2`, `r audBias_ci1`] and $\sigma^2_A=$ `r audVar_val` [`r audVar_ci1`, `r audVar_ci2`]. For the visual modality, we obtained $b_V=$ `r visBias_val` [`r visBias_ci2`, `r visBias_ci1`] and $\sigma^2_V=$ `r visVar_val` [`r visVar_ci1`, `r visVar_ci2`].

```{r unimodal, out.width = "\\textwidth", fig.pos = "!h", fig.cap = "Human responses in the unimodal conditions across the three experiments. Points represent the proportion of `different' to `same' responses in the auditory-only condition (left), and visual-only condition (right). Error bars are 95\\% confidence intervals. Solid lines represent best unimodal posterior fits."}

@@

correlation_plot <- ggplot(models_all, 
       aes(x = pred, y = joint, col = factor(concept_dist), 
           shape = factor(sound_dist))) + 
  geom_point(size =1)+
  scale_colour_solarized()+
 #geom_pointrange(aes(ymin = summary_ci_lower, ymax = summary_ci_upper), 
  #                position = position_dodge(width = .1), size=0.2) +

@@

#### Normative models
Figure\ \@ref(fig:bimodal) compares the predictions of the normative models against human responses. Remember that the normative models use variance estimates from the unimodal conditions (where people see input from only one modality) to predict data in the bimodal condition (where people see input from both modalities). Each point represents data form a particular audio-visual matching (e.g., the visual token $v$ whose distance from the target $v_t$ is $|v-v_t|=3$, matched with the auditory token $a$ whose distance from the target $a_t$ is $|a-a_t|=2$). The visual, auditory and optimal model explained, respectively, `r 100*R2_visual_1`%, `r 100*R2_auditory_1`%, and `r 100*R2_optimal_1`% of total variance in mean responses.

@@

```{r bimodal, out.width = "\\textwidth", fig.pos = "!h", fig.cap = "Human responses vs. Models' predictions in the bimodal condition across the three experiments. Each point represents data form a particular audio-visual matching. Shape represents auditory distance from the target, and color represents visual distance from the target. Thus, each point is characterized by both  shape and color."}

@@

## Discussion
Overall, we found that the optimal model explained much of the variance in the mean judgments, and largely more than what can be explained with the auditory or the visual models alone. Moreover, the high value of the coefficient of determination in the optimal model ($r^2$=`r R2_optimal_1`) suggests that the population was near-optimal. However, we see in Figure\ \@ref(fig:bimodal) that the mean responses deviated systematically from the optimal prediction in that they were slightly pulled toward chance (i.e., the probability 0.5). This fact is due to the increase in the value of the variance associated with each modality. Note however that, despite this increase in randomness, our analysis of modality preference showed that the relative values of these variances were not different (Figure\ \@ref(fig:bias)), meaning that there was no evidence for a modality preference. Thus, 1) There was a simultaneous increase in the values of the auditory and visual variances in the bimodal condition compared to the unimodal condition, meaning that the bimodal input lead to an increase in response randomness, and 2) this increased randomness did not affect the relative weighting of both modalities, i.e., the participants was weighting modalities according to the relative reliability predicted by the optimal model. This situation corresponds to the first case of sub-optimally described in Figure\ \@ref(fig:subOptim).

@@

```{r echo=FALSE, individual, out.width = "\\textwidth", fig.pos = "!h",fig.cap = "Distributions of individual values of the visual variance relative to the auditory variance in Experiment 1. Light color represents the real individual distribution, and dark color represents the simulated individual distribution sampled from the descriptive model."}

@@

In Experiment 1, we tested word recognition when there was multimodal uncertainty in terms of category membership only. In real life, however, tokens can undergo distortions due to noisy factors in the environment (e.g., car noise in the background, blurry vision in a foggy weather). In Experiment 2 and 3, we explore this additional level of uncertainty.
